draft

### 2025-05-21 draft.tex
text ready for shantanu

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution (that which comes from different phenomena from the training datasets) data \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

@@

By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility. We aim to make the morphological profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already extracted for multiple public datasets.

@@

The validation of cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of measurements from our implementation. A few data points fall outside the diagonals, which hint that some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm. This result provides evidence that cp\_measure can be confidently in cases where CellProfiler would be used, and the insights are unlikely to change.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocyte 3D nuclei data}
\label{sec:org447090b}

To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow. We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalinin3DCellNuclear2018}. We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells. We achieved 87\% classification accuracy in the testing dataset. Using the  we identified which features distinguish cells on the later samples and distinguish subpopulations. Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap}
\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.} Shapley values of the most important features to classify the day in which an image was taken (out of three). The test data accuracy is shown in bold.}
\end{figure}

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. This indicates that the major axis length of the cell to is an indicator of phenotypic effect over the course of the experiment, implying that cells became more elongated on their minor axis. Though unsurprising, as we expect astrocytes to extend over time, cp\_measure was able to uncover this with only a few lines of code.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these measurements as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological measurements yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figs/spatial.png}
\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}
\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they can present barriers for smooth adoption into computational workflows. In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). It enables simpler automated analyses of microscopy data in either short scripts and complex pipelines. It avoids graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.

@@

Already, cp\_measure  has been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. We therefore foresee cp\_measure being useful for a broad scientific community beyond morphological profiling.
\section{Future work}
\label{sec:org5cdbb12}
The most obvious way to make cp\_measure more useful is to contribute it back to CellProfiler. This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.

>>>>

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution (that which comes from different phenomena from the training datasets) data \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

@@

By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility. We aim to make the morphological profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent since they are freed from all the orchestration code, we reduced the amount of time and human effort required to integrate it into existing and new pipelines, while still retaining the standard features already extracted for multiple public datasets.

@@

The validation of cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}. We find that most features had a $R^2$ correlation larger than 90\%. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of measurements from our implementation. Some data points fall outside the diagonals for a small set of f, which hint that some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm. This result provides indicates that cp\_measure can be used in some cases instead of CellProfiler, and we would not expect a massive change in results.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocyte 3D nuclei data}
\label{sec:org447090b}

To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow. We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalinin3DCellNuclear2018}. We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells. We achieved 87\% classification accuracy in the testing dataset. Using the morphological profiles, we identified which features distinguish cells based on the time at which the image was taken. Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap}
\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.} Shapley values of the most important features to classify the day of differentiation based on the images (out of three). Each point is an image containing multiple nuclei. The test data accuracy is shown in bold. The x-axis indicates how impactful a given feature is on the classification accuracy. The y-axis are the most significant features, and the last entry is the rest of the features grouped.}
\end{figure}

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. This indicates that the minor axis length of the nuclei is an indicator of phenotypic effect over the course of the experiment, implying that nuclei becomes thicker (though not necessarily more round). Though unsurprising, as we expect astrocyte nuclei to extend over time, by using cp\_measure was able to uncover this with relatively few lines of code.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these measurements as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological measurements yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figs/spatial.png}
\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}
\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they can present barriers for smooth adoption into computational workflows. In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). Although there are still rough edges to polish, such as ensuring a 1:1 match, it enables simpler automated analyses of microscopy data in either short scripts and complex pipelines. It avoids graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.

@@

Already, cp\_measure  has been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. We therefore foresee cp\_measure being useful for a broad scientific community beyond morphological profiling.
\section{Future work}
\label{sec:org5cdbb12}
We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits: This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the original community for which it was originally developed.
